High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-18 doc/asplos2013/interface.tex
More writing for the asplos paper

The Legion Low-Level Runtime (LLR) interface organizes
functionality into objects.  Objects are grouped into
four different categories: machine objects for describing
the underlying architecture, events for capturing dependences,
deferred locks for performing synchronization, and physical regions
for managing data.  In all cases, the objects are light-weight
{\em handles} that can be passed around by value.  Furthermore
every handle is valid everywhere in the system.  This property allows 
a programmer to pass handles by value between computations without
having to reason about the distributed nature of the system.
We describe how we support this property in more detail in 
Section~\ref{sec:impl}.

\lstset{
  captionpos=b,
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  %morekeywords={region,coloring,partition,spawn,disjoint,aliased},
  %deletekeywords=float,
}
\subsection{Machine Interface}
\label{subsec:machmodel}
Listing~\ref{lst:machineapi} shows the machine interface.  The machine interface 
contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
the system there exists a unique {\tt Processor} handle naming that processor.  
{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
launch a {\em task} on that processor.  Note that because the {\tt spawn} operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  The {\tt spawn} call takes as arguments
the {\tt TaskFuncID} of the task to be run, a pointer to and size of arguments to
be passed to the task, and an optional {\tt Event} (described in Section~\ref{subsec:events})
that must trigger before the task can begin.  The result of the {\tt spawn} call
is an {\tt Event} that will trigger when the task completes.

For every memory in the system there also exists a unique {\tt Memory} handle. 
Different memories often, but not always, imply a different address space.  For
example, in a large cluster each node would have a different {\tt Memory} handle
corresponding to each node's memory.
However, if the cluster supported multiple GPUs then there would be a different
{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
address space with the corresponding node memory.  Memories will be used for describing
the placement of data described in Section~\ref{subsec:phyreg}.

The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the programmer
creates a {\tt Machine} object and invokes the {\tt run} method (line 30-31).
Second, the {\tt Machine} object provides an interface for the programmer to 
discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
sets of all the {\tt Processor} and all the {\tt Memory} handles in the system.
The {\tt Machine} also maintains two relations between these sets:

\begin{itemize}
\item Processor-Memory: for each {\tt Processor} and each {\tt Memory} whether
the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
latency and bandwidth properties if access is possible
\item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
performed between the two memories and latency 
bandwidth properties if a copy is possible
\end{itemize}

@@

In most cases events are created by the runtime as the result of other
operations and the runtime is responsible for triggering these events.  Users 
can create a special type of event called a {\tt UserEvent} (line 14) that the
user has the power to trigger.  {\tt UserEvent} is a sub-type of {\tt Event} 
that has {\tt trigger} method (line 17).  Users can also create another
type of event called a {\tt Barrier} that will require multiple arrivals before
the event is triggered.  The user can manage the number of arrivals that must
be seen as well as how many arrivals occur at a time.  Note that since {\tt UserEvent}
and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
{\tt Event} is required.

@@

>>>>

The runtime interface organizes
functionality into objects.  Objects are grouped into
four different categories: machine objects for describing
the underlying architecture, events for capturing dependences,
deferred locks for performing synchronization, and physical regions
for managing data.  In all cases, the objects are light-weight
{\em handles} that can be copied and passed around by value.  Furthermore
every handle is valid everywhere in the system.  This property allows 
a programmer to pass handles by value between computations without
having to reason about the distributed nature of the system.
We describe how we support this property in more detail in 
Section~\ref{sec:impl}.

\lstset{
  captionpos=b,
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  %morekeywords={region,coloring,partition,spawn,disjoint,aliased},
  %deletekeywords=float,
}
\subsection{Machine Interface}
\label{subsec:machmodel}
Listing~\ref{lst:machineapi} shows the machine interface.  The machine interface 
contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
the system there exists a {\tt Processor} handle with a unique ID naming that processor.  
{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
launch a {\em task} on that processor.  Note that because the {\tt spawn} operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  The {\tt spawn} call takes as arguments
the {\tt TaskFuncID} of the task to be run, a pointer to and size of arguments to
be passed to the task, and an optional {\tt Event} (described in Section~\ref{subsec:events})
that must trigger before the task can begin.  The result of the {\tt spawn} call
is an {\tt Event} that will trigger when the task completes.

For every memory in the system there also exists a {\tt Memory} handle with a unique ID. 
Different memories often, but not always, imply a different address space.  For
example, in a large cluster each node would have a different {\tt Memory} handle
corresponding to each node's DRAM memory.
However, if the cluster supported multiple GPUs then there would be a different
{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
address space with the corresponding node memory.  Memories will be used for describing
the placement of data described in Section~\ref{subsec:phyreg}.

The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the programmer
creates a {\tt Machine} object and invokes the {\tt run} method (line 30-31).
Second, the {\tt Machine} object provides an interface for the programmer to 
discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.
The {\tt Machine} also maintains two relations between these sets:

\begin{itemize}
\item Processor-Memory: for each {\tt Processor} - {\tt Memory} pair whether
the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
latency and bandwidth properties if access is possible
\item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
performed between the two memories and latency and
bandwidth properties if a copy is possible
\end{itemize}

@@

In most cases events are created by the runtime as the result of other
operations and the runtime is responsible for triggering these events.  Users 
can create a special type of event called a {\tt UserEvent} (line 14) that the
user has the power to trigger.  {\tt UserEvent} is a sub-type of {\tt Event} 
that has {\tt trigger} method (line 17).  Users can also create another
type of event called a {\tt Barrier} that will require multiple arrivals before
the event is triggered.  The user can manage the number of arrivals that must
be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
count goes to zero, the barrier will trigger.  Note that since {\tt UserEvent}
and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
{\tt Event} is required.

@@

{\em Physical regions} are the mechanism for reasoning about
the placement and movement of data in the runtime.  Physical regions are an 
allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
shows a subset of the physical region interface.  The first object in the interface
is a {\tt RegionMetaData} object (line 1).  {\tt RegionMetaData} objects provide the interface for the
creation of sets of physical regions that possess the same number and type of elements (but not necessarily
the same data layout).  The runtime is only able
to copy data between two physical regions that were created by the
same {\tt RegionMetaData} object.  A runtime error will be generated if a copy is attempted
between two physical regions that were not created by the same {\tt RegionMetaData}.

An instance of a physical region is represented by a {\tt RegionInstance} object (line 17).
When creating a physical region the programmer must specify the {\tt Memory}
in which the physical region is going to be allocated (line 10).  Once the physical region is created it
cannot be moved.   There is no coherence of data between physical regions created by the 
same {\tt RegionMetaData}.  It is the programmer's responsiblity to manage data coherence
using copies between physical regions (line 22-23).  Just like other operations,
copies can be predicated on an {\tt Event} and return an {\tt Event} corresponding to
completion.  A copy between two physical regions can only be performed if copies
are permitted between the two memories where the regions were created.  
The legality of copies can be discovered using queries to the {\tt Machine} object
described in Section~\ref{subsec:machmodel}.

The data held inside a physical region is accessed by another kind of object called an
{\em instance accessor} (not shown).  Accessors support read, write, and reduction operations
to elements inside the physical region.  Accessors can be specialized for the case where
all memory is known to be directly accessible to the executing processor, which allows 
for direct references to elements.  In cases where not all memory is directly accessable, 
accessors supply the level of indirection necessary to convert operations into remote 
memory operations (RMAs).  %An example of this is described in more detail in Section~\ref{sec:impl}.

One special (but common) case for applications is when reductions need to be performed.
For reductions it is usually more efficient to store reduction
operations in a local reduction buffer and then merge reduction buffers together at a later
point in time.  To support this case the interface allows for the creation of a special kind
of physical instance called a {\em reduction instance}.  The call to {\tt create\_instance} on
lines 13-14 takes a {\em reduction operation} to be associated with a physical instance which
tells the runtime to create a reduction instance.  Reduction instances can only be accessed
using a reduction of the given reduction type; any other accesses will result in a runtime
error.  A reduction operation is a functor which must have at least one method of the type
$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes an element of type $T_1$ and a value
to be reduced of type $T_2$ and creates and new element.  Reduction operations can also
support two additional methods: a method that supplies an initial value of type $T_2$ and a 
fold method of type $(T_2 \rightarrow T_2 \rightarrow T_2)$.  The presence of these methods
enable additional runtime optimizations for reductions described in Section~\ref{sec:impl}.

Reduction instances can be copied using the same interface as regular physical instances.
It is only legal to copy a reduction instance to a regular physical instance or another
reduction instance.  Any attempt to copy from a regular physical instance to a reduction
instance will result in a runtime error.  A copy from a reduction instance to a regular
physical instance will apply all the reductions in the reduction buffer to the physical
instance.  If a copy is from one reduction instance to another, the two reduction buffers
will be merged.  The semantics of a merge are dependent upon the underlying implementation
of reduction buffers which is discussed in more detail in Section~\ref{sec:impl}.

\begin{lstlisting}[float={t},label={lst:regionapi},caption={Subset of Physical Region Interface.}]
class RegionMetaData {
public:
  const unsigned id;
  static const RegionMetaData NO_REGION;
  // Create and destroy metadatas
  static RegionMetaData create_region(size_t num_elmts, 
                                      size_t elmt_size);
  void destroy_region() const;
  // Create and destroy instances
  RegionInstance create_instance(Memory memory) const;
  void destroy_instance(RegionInstance instance) const;
  // Create a physical instance only for reductions
  RegionInstance create_instance(Memory memory, 
                        ReductionOpID redopid) const;
};

class RegionInstance {
public:
  const unsigned id;
  static const RegionInstance NO_INST;
  // Copy between instances
  Event copy_to(RegionInstance target, 
                Event wait_on = Event::NO_EVENT);
};
\end{lstlisting}
